Random number generation and applications on GPU’s in R?

### 2021-10-05 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.Rnw
no important changes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@2889 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

The remaining sections are organized as follows. 
In Section 2, we introduce streams and the use of streams in work items on a GPU device, and describe and check the functions for generating uniform and Normal random numbers.
In Section 3, we apply GPU-generated uniform random numbers in Monte Carlo simulation for Fisher’s exact test, we parallelise this Monte Carlo simulation task and implement it on GPU, then we provide two real data examples to demonstrate the function usage and test its R performance.
Section 4 simulates simultaneously multiple Gaussian random surfaces by applying GPU-generated Normal random numbers and GPU-computed batches of Mat\'ern covariance matrices, then we plot the generated random surfaces in the end. 
Finally, the paper concludes with a short summary and a discussion in Section 5.

\end{leftbar}
<<echo=FALSE, results="hide", message=FALSE>>=
library("gpuR")
library("gpuRandom")
setContext(   grep('gpu', listContexts()$device_type) [1]    )
#setwd("/home/ruoyong/paper1_2021_5")
@

@@

\subsection{Creating streams}

{\tt changed createStreams functions to be compatible with each other}

create streams on cpu
<<createStreams, size="tiny">>=
myStreamsCpu <- createStreamsCpu(n=4, seed = 12345)
t(myStreamsCpu)
@

streams on gpu
<<createStreamsGpu>>=
myStreamsGpu1 = vclMatrix(myStreamsCpu)
myStreamsGpu2 = createStreams(n=4, seed=12345)
range(as.matrix(myStreamsGpu1 - myStreamsGpu2))
@

<<simStuff, eval=TRUE,tidy=TRUE>>=
sim_2 <-gpuRandom::runif(n=6, streams=myStreamsGpu1, Nglobal=c(1,nrow(myStreamsGpu1)), type = "double")
as.vector(sim_2)
@

@@

In \proglang{clRNG} library, a stream object is a structure that contains three states (see the first column in the \proglang{R} output below): the current state of the stream, the initial state (or seed) of the stream , and the initial state of the current substream. The function \fct{clrngMrg31k3pCreateStreams} from \proglang{clRNG} creates streams on the host computer using the MRG31k3p RNG. If we want to use the streams in work items on a GPU device, the streams have to be copied from the host to the global memory of the GPU device, and then each work item picks streams from the global memory and copies their current states only to its private memory. In \pkg{gpuRandom}, to avoid unnecessary data transfer between host and device, we are able to create streams both on the host computer and on the GPU device. \fct{gpuRandom::CreateStreamsCpu} is an \proglang{R} interface function of \fct{clrngMrg31k3pCreateStreams} that creates streams on the host as \proglang{R} matrices, so that users can easily view the streams. \fct{gpuRandom::CreateStreamsGpu} is adapted from \fct{clrngMrg31k3pCreateStreams} with the advantage of creating streams directly on GPU device. From our experience, transferring a large-size matrix of streams from host to device by converting it to a `\code{vclMatrix}' object can take very long time, hence, \fct{gpuRandom::CreateStreamsGpu} is more efficient and more frequently used. In the following, we demonstrate the usage of these two functions by generating 4 streams on the host and on the device respectively.
<<streams on CPU>>=
# Create streams on host
streams_on_CPU<-gpuRandom::CreateStreamsCpu(seed = rep(12345, 6), Nstreams=4)
t(streams_on_CPU)
@
\fct{gpuRandom::CreateStreamsCpu} has only one argument \code{Nstreams}, which is the number of streams to create. For the MRG31k3p RNG, each state of the stream object is comprised of six 31-bit integers, the state of the first stream (seed of the \proglang{clRNG} library) is by default ``12345''. By setting a seed same as the MRG31k3p default seed in \fct{gpuRandom::CreateStreamsGpu}, we can produce on device exactly the same streams as those created on host as follows,
<<streams on GPU1, eval=TRUE>>=
# Create streams on GPU
seed <- c(12345, 12345, 12345, 12345, 12345, 12345)
streams1 <- gpuRandom::CreateStreamsGpu(seedR=seed, Nstreams=4, keepInitial=TRUE)
streams_from_GPU <- as.matrix(streams1)
streams2 = deepcopy(streams1)
@ 
<<streams on GPU2, eval=TRUE, echo=FALSE>>=
gpuR::colnames(streams_from_GPU) = c("current.g1.1", "current.g1.2", "current.g1.3", "current.g2.1", "current.g2.2", "current.g2.3",
"initial.g1.1", "initial.g1.2", "initial.g1.3", "initial.g2.1", "initial.g2.2", "initial.g2.3")
#"substream.g1.1", "substream.g1.2", "substream.g1.3", "substream.g2.1", "substream.g2.2", "substream.g2.3")   
@ 
<<streams on GPU3,eval=TRUE>>=
t(streams_from_GPU)   
@
The \fct{gpuRandom::CreateStreamsGpu} has three arguments:
\begin{itemize}
  \item \code{seedR}: an R vector of length 6, which is the seed of streams. Default is set to a vector of six ``12345''.
  \item \code{Nstreams}: number of streams to create.
  \item \code{keepInitial}: logical specifying whether to keep the seed in generated streams. Default is set to TRUE.
\end{itemize}
\fct{CreateStreamsGpu} in \pkg{gpuRandom} are similar to \code{.Random.seed} in \proglang{R}. By setting the RNG status with the \fct{set.seed} function in \proglang{R}, we are able to repeat the sequence of random seeds, which is an important quality criteria of RNGs. Similarly in \pkg{gpuRandom}, with the \code{seedR} argument, we are able to produce exactly the same sequence of streams as many times as we want.

\subsection{Using streams to generate  i.i.d. $U (0, 1)$ random numbers on GPU}
The streams created on GPU are then used in work items that executes in parallel on GPU device. In \pkg{gpuRandom} each work item takes one distinct stream to generate random numbers, it is important to notice that the number of streams should always be equal to the number of total work items in use. The main part of the kernel (\proglang{OpenCL} functions for execution on the device) for generating uniform random numbers is shown in Listing \ref{lst:uniformkernel}. Kernels are written in the \proglang{OpenCL C} language, in which \code{__kernel} declares a function as a kernel, pointer kernel arguments must be declared with an address space qualifier, for example \code{__global} and \code{__local}. Here the pointers to \code{streams} and to the matrix \code{out} on the global memory are passed to the kernel as arguments. \code{Nrow} and \code{Ncol} represent number of rows and columns of the matrix \code{out} respectively. \code{NpadCol} is the internal number of columns of \code{out}, \code{NpadStreams} is internal number of columns of the \code{streams} matrix (these variables are defined in macros not shown here). Each stream's current state is copied to the private memory of each work item by the function \fct{streamsToPrivate}. \code{index} represents the position of the work item when mapping it to a one-dimensional data structure. The function \fct{clrngMrg31k3pNextState} generates a uniform random integer named \code{temp} between 0 and 4294967295,  the value of \code{temp} is then scaled to be in the interval $(0,1)$ by multiplying it by a constant \code{mrg31k3p\_NORM\_cl}, which is defined to be $1/4294967295$ in macro not shown here. At the end of generating random numbers, streams are transferred back to global memory through the function \fct{streamsFromPrivate}.
\begin{framed}
\begin{lstlisting}[caption={captiontext},language=C,basicstyle=\small,label={lst:uniformkernel}]
 __kernel void mrg31k3pMatrix(
    __global int* streams,
    __global double* out){

@@

    }//Dcol
 }//Drow
streamsFromPrivate(streams,g1,g2,startvalue);
}
\end{lstlisting}
\end{framed}
This kernel executes over a 2-dimensional index space. Figure \ref{fig:1} illustrates how this kernel executes in practice with a simple example. Suppose we have a $2 \times 4$ NDRange (index space size) ((a) in Figure \ref{fig:1}) and want to create a $4 \times 6$ matrix ((b) in Figure \ref{fig:1}) of uniform random numbers using 8 streams. In each grid of (a) that represents a work item, there are two rows of numbers, the number in the above row is the stream index (from 0 to 7), and a pair of numbers in the second row is the 2-dimensional global ID of the work item. Each stream object is passed to each work item. The number in a cell of the matrix in (b) corresponds to the stream, it indicates which stream does the random number generation for this matrix cell. For example, stream 0 which is allocated in work item $(0,0)$ generates a random number in matrix $(0,0)$ in the first iteration, then moves on to matrix $(0,2)$ and generates a random number there in the second iteration, in the third iteration it generates a random number in matrix $(0,4)$ and so on until it goes out of the matrix range in the next iteration.
\begin{figure}[H]%
    \centering
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/paper1_2021_5/f1} }}%
    \qquad
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/paper1_2021_5/f2} }}%
    \caption{name?\label{fig:1}}%
\end{figure}

Now we use the 4 streams created in Section 2.1 and an NDRange of 1 by 4 work items to generate a vector \code{sim_1} of 6 i.i.d. $U (0,1)$ random numbers using the function \fct{gpuRandom::runif}. If \code{streams} and \code{Nglobal} are both missing, then by default \code{Nglobal} is set as (64,8), and a `vclMatrix' of \code{prod(Nglobal)} streams is supplied. The output is an S4 object (a `\code{vclVector}' or a `\code{vclMatrix}'), to view the generated random numbers, we need to convert it to an R vector or matrix, in this way, the random numbers are moved from the global memory of device to the host.
<<eval=TRUE, tidy=TRUE>>=
sim_1 = gpuRandom::runif(n=6,streams=streams1, Nglobal=c(1,4), type="double")
as.vector(sim_1)
@
The arguments of the \fct{gpuRandom::runif} are described as follows:
\begin{itemize}
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or specifying the length if to create a vector.
  \item \code{streams}: streams used for random number generation.
  \item \code{Nglobal}: global index space that defines the total work-items that execute in parallel. Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' format of generated random numbers.
\end{itemize}
Next we show that by using a same stream \code{streams2} as \code{streams1},  \fct{gpuRandom::runif} returns a vector \code{sim_2} of same random numbers as \code{sim_1}. 
<<eval=TRUE,tidy=TRUE>>=
sim_2 <-gpuRandom::runif(n=6, streams=streams2, Nglobal=c(1,4), type = "double")
as.vector(sim_2)
@
Notice that \code{streams2} is a \code{deepcopy} of \code{streams1} before \code{streams1} is used. Reusing \code{streams1} will produce a vector different from \code{sim_1} (see \code{sim_3}). Because each time a random number is generated, the current state of the stream is changed (advances by one position), below shows the first six rows (current states) of \code{streams1} after it being used.
<<eval=TRUE,tidy=TRUE>>=
sim_3 = gpuRandom::runif(n=6,streams=streams1, Nglobal=c(1,4), type="double")
as.vector(sim_3)
t(as.matrix(streams1))[1:6,]
@
Unlike the objects created by \proglang{R}, by default they remain in the memory after a restart. Streams created on GPU does not remain in the memory when a new \proglang{R} session begins. However, we can save the streams on the CPU in a file using the \fct{save} function, and later recall the streams with the \fct{load} function, in this way we can reproduce the same sequence of random numbers in simulations. Below is a trivial example, in which we demonstrate saving \code{streams1} to a data file called \code{streams_from_GPU.rds} on CPU and then load it back and transfer it to GPU to produce the vector called \code{sim_saved}, which is also the same as \code{sim_1}.
<<echo=TRUE,eval=FALSE>>=
saveRDS(streams_from_GPU, "streams_from_GPU.rds")
@
<<save streams,eval=TRUE>>=
# Load the streams object as streams_saved
streams_saved <- gpuR::vclMatrix(readRDS("streams_from_GPU.rds"))
sim_saved = gpuRandom::runif(n=6,streams=streams_saved, 
                             Nglobal=c(1,4), type="double")
as.vector(sim_saved)
@

We evaluate the distribution of 50000 GPU-generated double-precision uniform random numbers by making a density histogram  (the left panel in Figure \ref{fig0}), and plotting its empirical cumulative distribution and overlaying the theoretical cumulative distribution (the right panel in Figure \ref{fig0}). The two cumulative distribution lines seem to overlap completely, suggesting the data generated by \fct{gpuRandom::runif} does come from a standard uniform distribution.
<<uniforms_random_vector,eval=TRUE>>=
streams <- CreateStreamsGpu(Nstreams =512, keepInitial=1)
random_vector<-gpuRandom::runif(n=50000, streams=streams, 
                                Nglobal=c(64,8), type="double")
random_vector<-as.vector(random_vector)
@
\begin{figure}[H]
\centering
<<largenumberuniforms, eval=TRUE, echo=FALSE, fig.height=4, fig.width=4, fig.show='hold'>>=
hist(random_vector, prob = TRUE, main="Histogram")
plot(ecdf(random_vector), ylab="F(x)",main="Empirical cumulative distribution")
curve(punif(x, 0, 1), lty=2, add=TRUE, col="red")
@
\caption{histogram and empirical cumulative distribution of the GPU-generated random numbers.\label{fig0}}
\end{figure}

@@

In the following code, We create a $1024 \times 512$ matrix named \code{normal_matrix} of single-precision floating-point Gaussian random numbers, and assess Normality by making a histogram and a Q-Q plot of the generated random numbers in Figure \ref{fig2}. The points in the Q-Q plot shown in the right panel of Figure \ref{fig2} fall along the diagonal line, we have good evidence the generated data are from Gaussian distribution. We find doing a Q-Q plot can occupy a large amount of time in the process of producing an R markdown file, so as an incidental work, we adapted the \proglang{R} function \fct{stat::qnorm} to make the function \fct{gpuRandom::qqnorm} that does the quantile computation for the normal distribution part on GPU, which can speed up the Q-Q plot greatly.
<<histAndQQ_prepare,eval=TRUE>>=
streams <- CreateStreamsGpu(Nstreams =512*128, keepInitial=1)
normal_matrix<-gpuRandom::rnorm(c(1024,512), streams=streams, 
                                Nglobal=c(512,128), type="float")
@
<<histAndQQ_prepare_for_cache, eval=TRUE, cache=TRUE>>=
avector<-as.vector(as.matrix(normal_matrix))
@
\begin{figure}[H]
\centering
 <<histAndQQ, echo=TRUE, eval=TRUE, cache=TRUE, fig.height=4, fig.width=4, fig.show='hold'>>=
hist(avector,breaks=40, main = "histogram")
gpuRandom::qqnorm(avector, Nglobal=c(128,64), main="Q-Q Plot")
@
\caption{Histogram and Q-Q plot of the GPU-generated random numbers.\label{fig2}}
\end{figure}

@@

\subsection{Comparing running time: Month data example}
<<Timecomparemonth,echo=FALSE,eval=TRUE, cache=FALSE, results="hide">>=
data2<-read.csv("/home/ruoyong/paper1_2021_5/month.csv") 
data2<-data2[,-1]
month<-as.matrix(data2)
rownames(month) <- c("Jan", "Feb","Mar","Apr","May","Jun","Jul","Aug","Sept","Oct","Nov","Dec")
colnames(month) <- c("Ane", "Men", "Cya", "Her", "Omp", "Gas", "Lim", "Cle", "Pal", "Down", "Chro", "Hypo")
Nglobal = c(128,128)
streams <- CreateStreamsGpu(Nstreams =128^2, keepInitial=1)
@
<<monthdata,eval=TRUE,echo=FALSE, cache=FALSE>>=
library(knitr)
library(kableExtra)
knitr::kable(month,format="latex",  caption = "Month\\label{tab:month}") %>%
  kable_styling(full_width = F, position = "center",latex_options = "HOLD_position")
@

<<timecomparemonth2Cpu, cache=TRUE, eval=TRUE>>=
#using CPU
stats::fisher.test(month,simulate.p.value = TRUE,B=1e6)$p.value
system.time(stats::fisher.test(month,simulate.p.value = TRUE,B=1e6))
@
\fct{stats::fisher.test} takes about 9.8 seconds (time varies a little bit from executions) for one million simulations, has a p-value of 0.4035.
<<timecomparemonth3Gpu, eval=TRUE,cache=FALSE>>=
almost.1 <- 1 + 64 * .Machine$double.eps
observed_m= -gpuRandom::logfactSum(month, c(16,16))/almost.1

month_GPU<-vclMatrix(month,type="integer")
result_monthgpu <- gpuRandom::fisher.sim(month_GPU, 1e6, streams=streams,
                   type="double", returnStatistics=TRUE,  Nglobal = c(128,128))
result_monthgpu$simNum
result_monthgpu$counts
result_monthgpu$p.value
@
We requested one million simulations on GPU while the actual number of simulation is 999424, and got 403274 cases whose test statitsic is below the observed threshold, so the p-value from \fct{gpuRandom::fisher.sim} for this table is about 0.40375, which is close enough to the p-value from \fct{stats::fisher.test}: 0.40380.  
<<timecomparemonth2Gpu, eval=TRUE,cache=FALSE>>=
#using GPU
system.time(gpuRandom::fisher.sim(month_GPU, 1e6, streams=streams, 
            type="double", returnStatistics=FALSE, Nglobal = c(128,128)))
@
\fct{gpuRandom::fisher.sim} takes about 0.32 seconds, \pkg{gpuRandom} accounts for 2.2\% of the elapsed time on CPU. 
<<timecomparemonth4Gpu, eval=TRUE,echo=FALSE, cache=TRUE>>=
result_monthgpu$cpu <- as.vector(result_monthgpu$sim)
@

\subsection{Comparing running time: Week data example}
<<Timecompareweek, echo=FALSE, eval=TRUE, cache=TRUE,results="hide">>=
library(latex2exp)
data<-read.csv("/home/ruoyong/paper1_2021_5/weekday.csv") 
#exampletwo<-as.matrix(as.data.frame(lapply(data, as.numeric)))
data<-data[,-1]
week<-as.matrix(data)
rownames(week) <- c("Mon", "Tue","Wed","Thu","Fri","Sat","Sun")
colnames(week) <- c("Ane", "Men", "Cya", "Her", "Omp", "Gas", "Lim", "Cle", "Pal", "Down", "Chro", "Hypo")
@

<<weekdata, eval=TRUE, echo=FALSE, cache=TRUE>>=
knitr::kable(week,format="latex",caption = "Week\\label{tab:week}") %>%
  kable_styling(full_width = F, position = "center",latex_options = "HOLD_position")
@
We perform 10 million simulations for the ``week'' table.
<<TimecompareweekCpu, cache=TRUE, eval=TRUE>>=
#using CPU
fisher.test(week,simulate.p.value = TRUE,B=1e7)$p.value
system.time(fisher.test(week,simulate.p.value = TRUE,B=1e7))
@
The ``week'' table has a much smaller p-value: around 0.0001208, which should require a larger number of simulations to get a more accurate p-value.
<<TimecompareweekGpu2, eval=TRUE, cache=FALSE>>=
observed_w= -gpuRandom::logfactSum(week, c(16,16))/almost.1

>>>>

The remaining sections are organized as follows. 
In Section 2, we introduce streams and the use of streams in work-items on GPU device, and describe and check the functions for generating uniform and Normal random numbers.
In Section 3, we apply GPU-generated uniform random numbers in Monte Carlo simulation for Fisher’s exact test, we parallelise this Monte Carlo simulation task and implement it on GPU, then we provide two real data examples to demonstrate the function usage and test its R performance.
Section 4 simulates simultaneously multiple Gaussian random surfaces by applying GPU-generated Normal random numbers and GPU-computed batches of Mat\'ern covariance matrices, then we plot the generated random surfaces in the end. 
Finally, the paper concludes with a short summary and a discussion in Section 5.

\end{leftbar}
<<echo=FALSE, results="hide", message=FALSE>>=
library("gpuR")
library("gpuRandom")
setContext(   grep('gpu', listContexts()$device_type) [1]    )
#setwd("/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5")
@

@@

\subsection{Creating streams}\label{create_streams}
In \proglang{clRNG} library, a stream object is a structure that contains three states: the current state of the stream, the initial state (or seed) of the stream , and the initial state of the current substream. The function \fct{clrngMrg31k3pCreateStreams} from \proglang{clRNG} creates streams on the host computer using the MRG31k3p RNG. If we want to use the streams in work-items on a GPU device, the streams have to be copied from the host to the global memory of the GPU device, and then each work-item picks streams from the global memory and copies their current states only to its private memory. \pkg{gpuRandom} is able to create streams both on the host computer and on the GPU device. \fct{createStreamsCpu} is an \proglang{R} interface function of \fct{clrngMrg31k3pCreateStreams} that creates streams on CPU as \proglang{R} matrices (see the \proglang{R} output below), in which the substreams are not created because they are not actually used in random number generation. \fct{createStreamsGpu} adapted the \fct{clrngMrg31k3pCreateStreams} and creates streams directly on GPU device, so that it can avoid unnecessary data transfer between host and device. From our experience, transferring a large-size matrix of streams from host to device by converting it to a `\code{vclMatrix}' object can take very long time, hence, \fct{createStreamsGpu} is more efficient and more frequently used. In the following, we demonstrate the usage of these two functions by generating 4 streams on the host and on the device respectively.
<<createStreams_CPU>>=
# creating streams on cpu
myStreamsCpu <- createStreamsCpu(n=4)
t(myStreamsCpu)
@
\fct{createStreamsCpu} has two arguments:
\begin{itemize}
  \item \code{n}: number of streams to create.
  \item \code{seed}: random seed of streams of length 6, recycled if shorter. Default is set to a vector of six ``12345''.
\end{itemize}
For the MRG31k3p RNG, each state of the stream is comprised of six 31-bit integers, the state of the first stream (seed of the \proglang{clRNG} library) is by default ``12345''. By setting a seed same as the MRG31k3p default seed in \fct{createStreamsGpu}, we can produce on device exactly the same streams as those created on host as follows,
<<createStreams_Gpu>>=
# creating streams on gpu
#myStreamsGpu1 = vclMatrix(myStreamsCpu)
myStreamsGpu1 = createStreamsGpu(n=4, seed=12345)
range(as.matrix(myStreamsGpu1) - myStreamsCpu)
matrix(as.matrix(myStreamsGpu1), nrow(myStreamsCpu), ncol(myStreamsCpu), dimnames = dimnames(myStreamsCpu))
@
\fct{createStreamsGpu} also has the two arguments \code{n}, which is the number of streams to create, and \code{seed}, which is the seed of streams, and is also by default a vector of six ``12345''. \fct{createStreamsGpu} is similar to \code{.Random.seed} in \proglang{R}. By setting the RNG status with the \proglang{R} function \fct{set.seed}, we are able to repeat the sequence of random seeds, which is an important quality criteria of RNGs. Similarly in \fct{createStreamsGpu} and \fct{createStreamsCpu}, with the \code{seed} argument, we are able to produce exactly the same sequence of streams as many times as we want.

@@

\subsection{Using streams to generate  i.i.d. $U (0, 1)$ random numbers on GPU}
The streams created on GPU are then used in work-items that executes in parallel on GPU device. In \pkg{gpuRandom} each work-item takes one distinct stream to generate random numbers, it is important to notice that the number of streams should always be equal to the number of total work-items in use. The main part of the kernel (\proglang{OpenCL} functions for execution on the device) for generating uniform random numbers is shown in Listing \ref{lst:uniformkernel}. Kernels are written in the \proglang{OpenCL C} language, in which \code{__kernel} declares a function as a kernel, pointer kernel arguments must be declared with an address space qualifier, for example \code{__global} and \code{__local}. Here the pointers to \code{streams} and to the matrix \code{out} on the global memory are passed to the kernel as arguments. \code{Nrow} and \code{Ncol} represent number of rows and columns of the matrix \code{out} respectively. \code{NpadCol} is the internal number of columns of \code{out}, \code{NpadStreams} is internal number of columns of the \code{streams} matrix (these variables are defined in macros not shown here). Each stream's current state is copied to the private memory of each work-item by the function \fct{streamsToPrivate}. \code{index} represents the position of the work-item when mapping it to a one-dimensional data structure. The function \fct{clrngMrg31k3pNextState} generates a uniform random integer named \code{temp} between 0 and 4294967295,  the value of \code{temp} is then scaled to be in the interval $(0,1)$ by multiplying it by a constant \code{mrg31k3p\_NORM\_cl}, which is defined to be $1/4294967295$ in macro not shown here. At the end of generating random numbers, streams are transferred back to global memory through the function \fct{streamsFromPrivate}.
\begin{framed}
\begin{lstlisting}[caption={captiontext},language=C,basicstyle=\small,label={lst:uniformkernel}]
 __kernel void mrg31k3pMatrix(
    __global int* streams,
    __global double* out){

@@

    }//Dcol
 }//Drow
streamsFromPrivate(streams,g1,g2,startvalue);
}
\end{lstlisting}
\end{framed}
This kernel executes over a 2-dimensional index space. Figure \ref{fig:1} illustrates how this kernel executes in practice with a simple example. Suppose we have a $2 \times 4$ NDRange ((a) in Figure \ref{fig:1}) and want to create a $4 \times 6$ matrix ((b) in Figure \ref{fig:1}) of uniform random numbers. In each grid of (a) that represents a work-item, there are two rows of numbers, the number in the above row is the stream index (from 0 to 7), and a pair of numbers in the second row is the 2-dimensional global ID of the work-item. Each stream object is passed to each work-item. The number in a cell of the matrix corresponds to the stream index, it indicates which stream does the random number generation for this matrix cell. For example, stream 0 which is allocated in work-item $(0,0)$ generates a random number in matrix $(0,0)$ in the first iteration, then moves on to matrix $(0,2)$ and generates a random number there in the second iteration, in the third iteration it generates a random number in matrix $(0,4)$ and so on until it goes out of the matrix range in the next iteration.
\begin{figure}[H]%
    \centering
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/paper1_2021_5/f1} }}%
    \qquad
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/paper1_2021_5/f2} }}%
    \caption{name?\label{fig:1}}%
\end{figure}

Now we use the 4 streams created in section \ref{create_streams} and an NDRange of 1 by 4 work-items to generate a vector \code{sim_1} of 6 i.i.d. $U (0,1)$ random numbers using the function \fct{gpuRandom::runif}. If \code{streams} and \code{Nglobal} are both missing, then by default \code{Nglobal} is set as (64,8), and a `vclMatrix' of \code{prod(Nglobal)} streams is supplied. The output is an S4 object (a `\code{vclVector}' or a `\code{vclMatrix}'), to view the generated random numbers, we need to convert it to an R vector or matrix, in this way, the random numbers are moved from the global memory of device to the host.
<<eval=TRUE, tidy=TRUE>>=
myStreamsGpu2 = deepcopy(myStreamsGpu1)
sim_1 = runif(n=6,streams=myStreamsGpu1, Nglobal=c(1,4), type="double")
as.vector(sim_1)
@
The arguments of the \fct{gpuRandom::runif} are described as follows:
\begin{itemize}
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams used for random number generation.
  \item \code{Nglobal}: global index space (NDRange). Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' format of generated random numbers.
\end{itemize}
Next we show that by using a stream \code{myStreamsGpu2} same as \code{myStreamsGpu1},  \fct{runif} returns a vector \code{sim_2} of same random numbers as \code{sim_1}. 
<<eval=TRUE,tidy=TRUE>>=
sim_2 <-runif(n=6, streams=myStreamsGpu2, Nglobal=c(1,4), type = "double")
as.vector(sim_2)
@
Notice that \code{myStreamsGpu2} is a \code{deepcopy} of \code{myStreamsGpu1} before \code{myStreamsGpu1} is used. Reusing \code{myStreamsGpu1} will produce a vector different from \code{sim_1} (see \code{sim_3}). Because each time a random number is generated, the current state of the stream is changed (advances by one position), below shows the first six rows (current states) of \code{myStreamsGpu1} after it being used.
<<eval=TRUE,tidy=TRUE>>=
sim_3 = runif(n=6,streams=myStreamsGpu1, Nglobal=c(1,4), type="double")
as.vector(sim_3)
t(as.matrix(myStreamsGpu1))[1:6,]
@
Unlike the objects created by \proglang{R}, by default they remain in the memory after a restart. Streams created on GPU does not remain in the memory when a new \proglang{R} session begins. However, we can save the streams on the CPU in a file using the \fct{save} function, and later recall the streams with the \fct{load} function, in this way we can reproduce the same sequence of random numbers in simulations. Below is a trivial example, in which we demonstrate saving \code{streams1} to a data file called \code{streams_from_GPU.rds} on CPU and then load it back and transfer it to GPU to produce the vector called \code{sim_saved}, which is also the same as \code{sim_1}.
<<save streams,echo=TRUE,eval=TRUE>>=
saveRDS(as.matrix(createStreamsGpu(n=4)), "myStreams.rds")
# Load the streams object as streams_saved
streams_saved <- vclMatrix(readRDS("myStreams.rds"))
sim_saved = runif(n=6,streams=streams_saved, Nglobal=c(1,4), type="double")
as.vector(sim_saved)
@

We evaluate the distribution of 50000 GPU-generated double-precision uniform random numbers by making a density histogram  (the left panel in Figure \ref{fig0}), and plotting its empirical cumulative distribution and overlaying the theoretical cumulative distribution (the right panel in Figure \ref{fig0}). The two cumulative distribution lines seem to overlap completely, suggesting the data generated by \fct{runif} does come from a standard uniform distribution.
<<uniforms_random_vector,eval=TRUE>>=
streams <- createStreamsGpu(n = 512)
random_u<-runif(n=50000, streams=streams, Nglobal=c(64,8), type="double")
random_u<-as.vector(random_u)
@
\begin{figure}[H]
\centering
<<largenumberuniforms, eval=TRUE, echo=FALSE, fig.height=4, fig.width=4, fig.show='hold'>>=
hist(random_u, prob = TRUE, main="Histogram")
plot(ecdf(random_u), ylab="F(x)",main="Empirical cumulative distribution")
curve(punif(x, 0, 1), lty=2, add=TRUE, col="red")
@
\caption{Histogram and empirical cumulative distribution of the GPU-generated random numbers.\label{fig0}}
\end{figure}

@@

In the following code, We create a $1024 \times 512$ matrix named \code{normal_matrix} of single-precision floating-point Gaussian random numbers, and assess Normality by making a histogram and a Q-Q plot of the generated random numbers in Figure \ref{fig2}. The points in the Q-Q plot shown in the right panel of Figure \ref{fig2} fall along the diagonal line, we have good evidence the generated data are from Gaussian distribution. We find doing a Q-Q plot can occupy a large amount of time in the process of producing an R markdown file, so as an incidental work, we adapted the \proglang{R} function \fct{stat::qnorm} to make the function \fct{gpuRandom::qqnorm} that does the quantile computation for the normal distribution part on GPU, which can speed up the Q-Q plot greatly.
<<histAndQQ_prepare,eval=TRUE,cache=TRUE>>=
streams <- createStreamsGpu(n =512*128)
normal_matrix<-gpuRandom::rnorm(c(1024,512), streams=streams, 
                                Nglobal=c(512,128), type="float")
avector<-as.vector(as.matrix(normal_matrix))
@
\begin{figure}[H]
\centering
 <<histAndQQ, echo=TRUE, eval=TRUE, cache=TRUE, fig.height=4, fig.width=4, fig.show='hold'>>=
hist(avector,breaks=40, main = "histogram")
gpuRandom::qqnorm(avector, Nglobal=c(128,64), main="Q-Q Plot")
@
\caption{Histogram and Q-Q plot of the GPU-generated random numbers.\label{fig2}}
\end{figure}

@@

\subsection{Comparing running time: Month data example}
<<Timecomparemonth,echo=FALSE,eval=TRUE, cache=FALSE, results="hide">>=
data2<-read.csv("/home/ruoyong/paper1_2021_5/month.csv") 
data2<-data2[,-1]
month<-as.matrix(data2)
rownames(month) <- c("Jan", "Feb","Mar","Apr","May","Jun","Jul","Aug","Sept","Oct","Nov","Dec")
colnames(month) <- c("Ane", "Men", "Cya", "Her", "Omp", "Gas", "Lim", "Cle", "Pal", "Down", "Chro", "Hypo")
Nglobal = c(256,128)
streams <- createStreamsGpu(n =256*128)
@
<<monthdata,eval=TRUE,echo=FALSE, cache=FALSE>>=
library(knitr)
library(kableExtra)
knitr::kable(month,format="latex",  caption = "Month\\label{tab:month}") %>%
  kable_styling(full_width = F, position = "center",latex_options = "HOLD_position")
@

<<timecomparemonth3Gpu, eval=TRUE,cache=FALSE>>=
almost.1 <- 1 + 64 * .Machine$double.eps
observed_m= -logfactSum(month, c(16,16))/almost.1

month_gpu<-vclMatrix(month,type="integer")
result_month <- fisher.sim(month_gpu, 1e6, streams=streams,
                   type="double", returnStatistics=TRUE,  Nglobal = c(256,128))
result_month$simNum
result_month$counts
result_month$p.value
@
We requested one million simulations on GPU while the actual number of simulation is 1015808, and got 403274 cases whose test statitsic is below the observed threshold, so the p-value from \fct{gpuRandom::fisher.sim} for this table is about 0.40375, which is close enough to the p-value from \fct{stats::fisher.test}: 0.40380.  
<<timecomparemonth2Gpu, eval=TRUE,cache=FALSE>>=
#using GPU
system.time(fisher.sim(month_gpu, 1e6, streams=streams, 
            type="double", returnStatistics=FALSE, Nglobal = c(256,128)))
@
\fct{gpuRandom::fisher.sim} takes about 0.32 seconds, \pkg{gpuRandom} accounts for 2.2\% of the elapsed time on CPU. 
<<timecomparemonth4Gpu, eval=TRUE,echo=FALSE, cache=TRUE>>=
month_stats <- as.vector(result_month$sim)
@

<<timecomparemonth2Cpu, cache=TRUE, eval=TRUE>>=
#using CPU
stats::fisher.test(month,simulate.p.value = TRUE,B=1015808)$p.value
system.time(stats::fisher.test(month,simulate.p.value = TRUE,B=1015808))
@
\fct{stats::fisher.test} takes about 9.8 seconds (time varies a little bit from executions) for 1015808 simulations, has a p-value of 0.4035.

\subsection{Comparing running time: Week data example}
<<Timecompareweek, echo=FALSE, eval=TRUE, cache=TRUE,results="hide">>=
library(latex2exp)
data<-read.csv("/home/ruoyong/paper1_2021_5/weekday.csv") 
#exampletwo<-as.matrix(as.data.frame(lapply(data, as.numeric)))
data<-data[,-1]
week<-as.matrix(data)
rownames(week) <- c("Mon", "Tue","Wed","Thu","Fri","Sat","Sun")
colnames(week) <- c("Ane", "Men", "Cya", "Her", "Omp", "Gas", "Lim", "Cle", "Pal", "Down", "Chro", "Hypo")
@
<<weekdata, eval=TRUE, echo=FALSE, cache=TRUE>>=
knitr::kable(week,format="latex",caption = "Week\\label{tab:week}") %>%
  kable_styling(full_width = F, position = "center",latex_options = "HOLD_position")
@
